\documentclass{beamer}
\usetheme{Boadilla}

\usepackage{tikz}
\usepackage{amsmath}
\usepackage{unicode-math}
\usepackage{mathtools}
\usepackage[conor]{agda}

\setmathfont{XITS Math}
\input{latex/snippets} % agda code snippets

% manual agdfa typesetting
\newcommand{\xvar}{\AgdaBound{$x$}}
\newcommand{\yvar}{\AgdaBound{$y$}}
\newcommand{\nil}{\AgdaField{$[]$}}
\newcommand{\cons}{\AgdaField{$\dblcolon$}}
\newcommand{\phitm}{\AgdaFunction{$\phi$}}
\newcommand{\datatm}{\AgdaDatatype{Tm}}
\newcommand{\abstm}{\AgdaField{abs}}
\newcommand{\union}{\AgdaFunction{$\cup$}}

\title[Extenensional Fin. Sets \& Multisets in TT]{Extensional Finite Sets and Multisets in Type Theory}
% \subtitle{Free Algebraic Structures via Fresh Lists}
\author[S. Watters]{Clemens Kupke \and Fredrik Nordvall Forsberg \and \underline{\textbf{Sean Watters}}}
\institute{University of Strathclyde}
\date{05/04/2024}

\begin{document}


\begin{frame}
  \titlepage{}
\end{frame}


\begin{frame}{Scottish Programming Languages and Verification Summer School 2024}
\begin{itemize}
  \item What: Topics in TCS, particularly foundations of PL \& formal verification.
  \item Where: \textbf{University of Strathclyde, Glasgow}
  \item When: \textbf{Monday 29 July -- Friday 2 August}.

  \pause

  \item Invited Lecturer: Lindsey Kuper, UC Santa Cruz.
  \item Core courses on Category Theory and Type Theory by Chris Heunen and Conor McBride.
  \item Specialised courses by Bob Atkey, Kathrin Stark, Vikraman Choudhury, Andr\'{e}s Goens, Sam Lindley, and Jules Hedges.
\end{itemize}

\begin{center}
  \url{https://scottish-pl-institute.github.io/splv/2024-strathclyde}
\end{center}
\end{frame}


\begin{frame}{Type Theory 101: Types!}
  \begin{block}{Simple Data Types}
  \snippetdatalist{}
  \end{block}
\pause
  \begin{block}{Dependent Type Families}
  \snippetmember{}
  \end{block}
  %data types, identity type, funext, UIP/K, prop universe
\end{frame}

\begin{frame}{Type Theory 101: The Identity Type}
  \snippetidentity{}
  \pause
  \begin{block}{Universes}
    \begin{itemize}
    \item Types live in \emph{universes}. The universe of all (small) types is usually called $\AgdaDatatype{Type}$.
    \pause
    \item A \emph{proposition} is a type where $\forall x,y.~x \equiv y$.
    \pause
    \item A \emph{set} is a type whose identity type is a proposition.
    \end{itemize}
  \end{block}

  \pause
  \begin{block}{Function Extensionality (consistent but not provable in \emph{this} TT)}
    For any functions $f, g : A \to B$,
    \begin{center}
    $f \equiv g$ iff $\forall x. f~x \equiv g~x$.
    \end{center}
  \end{block}
\end{frame}

\begin{frame}{Motivation: A Normal Form for Contexts}
\begin{exampleblock}{A Syntax with Binding}
\snippetdatatm{}
\end{exampleblock}
\pause

\begin{itemize}
  \item  Consider the term $\abstm~\xvar~(\abstm~\yvar~\AgdaHole{?})$.
        \pause

  \item  Notice that the hole has type $\datatm~(\yvar~\cons~\xvar~\cons~\nil)$.
        \pause

\item  This means if we had some $\phitm : \datatm~(\xvar~\cons~\yvar~\cons~\nil)$, it would not fit!
\end{itemize}
\pause

\begin{block}{Goal}
  If we need such a $\phi$ to fit in such a hole, then we need contexts that are \emph{sets}, not lists.
\end{block}
\end{frame}


\begin{frame}{Sets, Subsets and Multisets in Type Theory}
\begin{block}{In Short}
\begin{itemize}
  \item Sets are (a certain class of) \emph{types}.
        \pause
  \item Given some $S : \mathsf{Set}$, subsets of $S$ are unary predicates $S \to \mathsf{Prop}$.
        \pause
  \item \emph{Decidable} subsets are functions $S \to 2$.
        \pause
\item Multisets over $S$ are functions $S \to \mathbb{N}$.
\end{itemize}
\end{block}
\pause

\begin{center}
But what about \emph{finite} sets/multisets?
\end{center}
\end{frame}

\begin{frame}{Finiteness, Decidable Equality, Extensionality}
\begin{block}{How to enforce finiteness?}
\begin{itemize}
  \item An enumeration list? \\
        (No extensionality.)
        \pause
  \item Bijection with $\AgdaDatatype{Fin}~n$, for some $n : \mathbb{N}$? \\
        (No extensionality, and we only get decidable equality with function extensionality.)
        \pause
  \item The \emph{mere existence} (propositional trucation) of such an enumeration? \\
        (No decidable equality.)
        \pause
  \item A higher-inductive type? \\
        (Works, but restricts us to HoTT. Also, HITs can be challenging. cf. Choudhury \& Fiore, 2023.)
\end{itemize}
\pause
\end{block}

\begin{center}
Our approach: an enumeration, but using a \emph{sorted-by-construction} list.
\end{center}
\end{frame}


\begin{frame}{Towards Free Commutative Algebraic Structures}
\begin{block}{Some Facts}
\begin{itemize}
  \item The free monoid over a set is given by the list functor, $\AgdaDatatype{List} : \AgdaDatatype{Set} \to \AgdaDatatype{Mon}$.
        \pause
  \item The free commutative monoid over a set is given by the finite multiset (AKA ``bag'')
        functor, $\AgdaDatatype{Bag} : \AgdaDatatype{Set} \to \AgdaDatatype{CMon}$.
        \pause
  \item The free idempotent commutative monoid over a set is given by the finite powerset
        functor, $\AgdaDatatype{\ensuremath{\mathcal{P}_{f}}} : \AgdaDatatype{Set} \to \AgdaDatatype{ICMon}$.
\end{itemize}
\end{block}
\pause

\begin{center}
This gives us a way to evaluate the correctness of our proposed solutions for sets and multisets; they should satisfy the above universal properties.
\end{center}
\end{frame}


\begin{frame}{Fresh Lists}
  \snippetdatafreshlist{}

  Originally due to Catarina Coquand.
  Generalisation to an arbitrary $R$ due to the Agda standard library.
\end{frame}


\begin{frame}{Sorted Lists}

\begin{itemize}
  \item To instantiate fresh lists as sorted lists, we set the relation to some strict total order, $<$.

  \item The ordering ensures that any two lists with the same members are equal (extensionality!).

  \item Strictness forces any given element to appear exactly once in any given list.
\end{itemize}

  The key definition is $\union : \AgdaDatatype{FList}(X,<) \to \AgdaDatatype{FList}(X,<) \to \AgdaDatatype{FList}(X,<)$, which merges two lists:

  \pause
  \begin{block}{Defintion: Merge for Sorted Lists}
    Redacted! (this definition is exactly what you'd expect, but with the added noise of the proof data)
  \end{block}
\end{frame}


\begin{frame}{Extensionality Principle}
\begin{itemize}
  \item We need to show that \union~satisfies the idem. comm. monoid laws.
  \item I redacted the definition of \union, so imagine how awful these proofs must be!
\end{itemize}

\begin{center}
  Except\ldots{}
\end{center}
\pause

\begin{block}{The Extensionality Principle for Sorted Lists}
  For all $xs, ys : \AgdaDatatype{FList}(X,<)$:

  \begin{center}
    $xs = ys$ iff $(a \in xs) \iff (a \in ys)$ for all $a : X$.
  \end{center}
\end{block}

This follows by appeal to decidability.
\end{frame}


\begin{frame}{The Categories at Issue}
\begin{itemize}
  \item We need ordering data to form the type $\AgdaDatatype{FList}(X,<)$.
  \item This needs to be preserved by round trips.
  \item So the objects of \emph{both} categories (i.e. the sets and the monoids) need to carry it.
\end{itemize}
\pause

\begin{block}{The Category \AgdaDatatype{STO}}
\begin{itemize}
  \item Objects: Sets, equipped with strict total orders.
  \item Morphisms: \emph{Not-necessarily monotone} functions on the underlying sets.
\end{itemize}
\end{block}

\begin{block}{The Category \AgdaDatatype{OICMon}}
\begin{itemize}
  \item Objects: Idempotent commutative monoids, equipped with strict total orders.
  \item Morphisms: Morphisms of idem. comm. monoids. i.e., functions on the underlying sets that preserve multiplication and identity, but \emph{not necessarily the ordering}.
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{Why No Monotonicity?}

Two reasons:
\begin{itemize}
\item We don't \emph{want} it --- we'd rather map arbitrary functions and have the output be sorted.
\item It breaks the adjunction (I think).
\end{itemize}

\pause

\begin{block}{Ignorance is Classically Valid, but Constructive Taboo}
 If only there was a ``free strict total order on a set'', then we could ignore the ordering data and obtain the genuine $\mathcal{P}_{f}$.
 But such a thing is a weak form of AC, and implies LEM.
\end{block}
\end{frame}

\begin{frame}{A Free-Forgetful Adjunction}
\begin{block}{The Universal Property}
   $\AgdaDatatype{SList} : \AgdaDatatype{STO} \to \AgdaDatatype{OICMon}$ forms a functor which is left adjoint to the forgetful functor $\AgdaDatatype{\ensuremath{\mathcal{u}}} : \AgdaDatatype{OICMon} \to \AgdaDatatype{STO}$ defined by $\AgdaDatatype{\ensuremath{\mathcal{u}}}(X,<,\cdot,\epsilon) \coloneq (X,<)$.
 \end{block}
 \pause
We prove the adjunction via natural bijection of homsets. That is, we must show:

\begin{center}
  $\mathsf{Hom}_{OICMon}(\AgdaDatatype{FList}(A),~B) \cong \mathsf{Hom}_{STO}(A,~ \AgdaDatatype{\ensuremath{\mathcal{u}}}(B))$
\end{center}
\pause

The hard direction involves showing that
\begin{center}
$\lambda~(f : \AgdaField{Carrier}~A \to \AgdaField{Carrier}~B)~\to~(\AgdaFunction{foldr}~(\lambda~a~b~\to~(f~a) \cdot b)~\epsilon)$
\end{center}
is a morphism of \AgdaDatatype(OICMon).


\end{frame}

\begin{frame}{Multisets\ldots{}}

  \ldots{} follow exactly the same pattern! Well, almost:

  \begin{block}{Extensionality Principle for $\AgdaDatatype{FList}(X, \leq)$}
    For all $a : X$, and $xs, ys : \AgdaDatatype{FList}(X, \leq)$:
  \begin{itemize}
    \item Each $\AgdaDatatype{FList}(X, \leq)$ induces a ``multiplicity function'', $\AgdaFunction{count} : \AgdaDatatype{FList}(X,\leq) \to X \to \mathbb{N}$, such that \\
          $\AgdaFunction{count}~xs~a~=~\AgdaFunction{count}~ys~a$ iff $(a \in xs) \cong (a \in ys) $.
    \item $(a \in xs) \cong (a \in ys)$ iff $xs = ys$.
  \end{itemize}
 \end{block}

 Then it's pretty much the same after that. We end up with:

 \begin{block}{Universal Property of Ordered Commutative Monoids}
   $\AgdaDatatype{SListD} : \AgdaDatatype{DTO} \to \AgdaDatatype{OCMon}$ forms a functor which is left adjoint to the forgetful functor $\AgdaDatatype{\ensuremath{\mathcal{u}}} : \AgdaDatatype{OCMon} \to \AgdaDatatype{DTO}$.
 \end{block}

 Where DTO and OCMon are analogous to STO and OICMon from before, but without strictness or idempotency.
\end{frame}


\begin{frame}{Other Results}
\begin{center}
\begin{tabular}{  |c|m{7em}|m{12em}| }
  \hline
  Freshness Relation & \centering{Data Structure} & \centering{Free Algebraic Structure} \tabularnewline
  \hline
  $\leq$, a total order & Sorted lists & Ordered comm. monoid \\
  $<$, a strict total order & Sorted lists w/o duplicates & Ordered idempotent comm.\ monoid \\
  $\lambda x. \lambda y. \bot$ & \AgdaDatatype{Maybe} & Pointed set \\
  $\lambda x. \lambda y. \top$ & \AgdaDatatype{List} & Monoid \\
  $\neq$ & Lists w/o duplicates & Left-regular band monoid \\
  $=$ & $1 + (A \times \mathbb{N}^{>0})$ & Reflexive partial monoid \\
  \hline
\end{tabular}
\end{center}
\end{frame}


\begin{frame}{References}
\begin{itemize}
  \item Kupke, C., Nordvall Forsberg, F., Watters, S.: \emph{A fresh look at commutativity: free algebraic structures via fresh lists}. In: APLAS '23. \\
        \url{https://doi.org/10.1007/978-981-99-8311-7_7}
  \item Our Agda formalisation: \url{https://seanwatters.uk/agda/fresh-lists}
  \item Coquand, C.: \emph{A formalised proof of the soundness and completeness of a simply
        typed lambda-calculus with explicit substitutions}. In: Higher Order Symbolic Computation, 2002. \\
        \url{https://doi.org/10.1023/A:1019964114625}
  \item Choudhury, V., Fiore, M.: \emph{Free commutative monoids in Homotopy Type Theory}. In: MFPS '22. \\
        \url{https://doi.org/10.46298/entics.10492}
\end{itemize}
\end{frame}

\end{document}
